---
title: "Network Models: Modeling Pairwise TCR Relationships"
author: "GW McElfresh"
date: "`r Sys.Date()`"
output: 
  rmdformats::readthedown:
    toc_depth: 5
vignette: >
  %\VignetteIndexEntry{Network Models: Modeling Pairwise TCR Relationships}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6
)
```

## Overview

`ModelTCRDyads()` fits two generalized linear models over all unique clone pairs
(dyads) derived from a TCR distance matrix and returns marginal estimates
describing pairwise relationships between levels of a grouping variable
(e.g. treatment arm or phenotype).

- **Connectivity model** (binomial): predicts the probability that two clones
  are connected for each dyad type. When `communityMethod = "DIANA"`,
  DIANA clustering is run internally at `distanceThreshold` (used as the
  dendrogram cut height). When `communityMethod = "threshold"`, a pair is
  connected if its distance is ≤ `distanceThreshold`.
- **Distance model** (Poisson): predicts the mean tcrdist distance among
  connected pairs for each dyad type.

By default (`estimand = "effects"`), marginal effects are estimated via
`avg_comparisons()`, returning pairwise contrasts between `pair_type` levels.
Setting `estimand = "means"` returns marginal means (predicted values per
level) via `avg_predictions()` instead.

Dyad types are represented by `pair_type`, an unordered factor whose levels are
lexicographically sorted group-pair labels (e.g. `"Controller:Controller"`,
`"Controller:Progressor"`). Because the relationship is symmetric
(`G1:G2 == G2:G1`), each unique pairing has exactly one level.

## Setup

```{r libraries, message=FALSE, warning=FALSE}
library(tcrClustR)
library(Seurat)
```

## Load example data

```{r loadData}
data_path <- system.file("extdata", "small_RIRA.rds", package = "tcrClustR")
seuratObj <- readRDS(data_path)
```

## Create Experimental Metadata

```{r}
#drop clones with unresolved subject Ids
seuratObj <- subset(seuratObj, subset = !is.na(SubjectId))

#assign subjects to phenotypes (this is just an example, real datasets should have this already)
protectedSubjects <- sample(unique(seuratObj$SubjectId), 
                            size = floor(length(unique(seuratObj$SubjectId)) / 2),
                            replace = FALSE)
seuratObj$Phenotype <- ifelse(seuratObj$SubjectId %in% protectedSubjects, "Protected", "NotProtected")
```


## Compute distances

`ModelTCRDyads()` requires distance matrices from `CalculateTcrDistances()`.
No prior call to `RunTcrClustering()` is needed — when
`communityMethod = "DIANA"` (the default), DIANA clustering is run internally.

```{r computeDistances}
seuratObj_TCR <- CalculateTcrDistances(
  inputData        = seuratObj,
  chains           = "TRB",
  minimumCloneSize = 2
)
```




## Fit dyad models

### DIANA community method (default)

The DIANA path runs divisive hierarchical clustering internally on the
distance matrix, using `distanceThreshold` as the dendrogram cut height
(analogous to `dianaHeight` in `RunTcrClustering()`). Two clones are
"connected" if they belong to the same DIANA cluster.

**NOTE**: The distance threshold used in this vignette is too permissive to be biologically reasonable — it is set high to ensure we have enough connections in this small example dataset. Real datasets will likely require a more stringent threshold (e.g. 20-50 for TRB CDR3 amino acid distances) and may require tuning to find a good balance between connectivity and sparsity.

```{r modelDiana}
result_diana <- ModelTCRDyads(
  seuratObj_TCR        = seuratObj_TCR,
  chains               = "TRB",
  groupColumn          = "Phenotype",
  estimand = 'effects', 
  distanceThreshold    = 100,
  communityMethod      = "DIANA",
  clusterSizeThreshold = 2,
  subjectIdCol         = "SubjectId",
  verbose              = TRUE
)
```

### Threshold community method

The threshold path treats `distanceThreshold` as a raw distance cutoff — a
pair is connected if its tcrdist distance is ≤ the threshold. Louvain
community detection is used for community annotation.

```{r modelThreshold}
result_threshold <- ModelTCRDyads(
  seuratObj_TCR     = seuratObj_TCR,
  chains            = "TRB",
  groupColumn       = "Phenotype",
  distanceThreshold = 100,
  communityMethod   = "threshold",
  subjectIdCol      = "SubjectId"
)
```

## Interpret results

`ModelTCRDyads()` returns a list with four elements:

- `connectivity` — marginal estimates for the connectivity model
- `distance` — marginal estimates for the distance model
- `models` — the fitted model objects and data
- `unpaired_summary` — per-clone connection partner summary

### Connectivity estimates

```{r connectivity}
print(result_diana$connectivity)
```

### Distance estimates

```{r distance}
print(result_diana$distance)
```

### Unpaired summary

A per-clone table showing which groups each clone connects to:

```{r unpaired}
head(result_diana$unpaired_summary)
```


### Plot model estimates

`PlotDyadEstimates()` generates a forest plot directly from the
`ModelTCRDyads()` output. It detects whether the result contains marginal
effects or marginal means and adjusts the plot accordingly. When plotting
connectivity, axis labels are tailored to emphasize that the estimates
represent probabilities.

```{r plotConnectivity}
PlotDyadEstimates(result_diana, which = "connectivity")
```

```{r plotDistance}
PlotDyadEstimates(result_diana, 
                  which = "distance", 
                  )
```

### Unpaired clone rates

Before subsetting to connected pairs for the distance model,
`ModelTCRDyads()` computes the per-group probability that a clone is
**unpaired** (i.e., has no connections to any other clone). These rates are
stored in `result$unpaired_rates` with exact binomial 95% confidence
intervals.

```{r unpairedRates}
print(result_diana$unpaired_rates)
```

Setting `showUnpaired = TRUE` in `PlotDyadEstimates()` adds a second facet
to the connectivity plot showing these unpaired probabilities alongside the
model-based connectivity estimates:

```{r plotUnpaired}
PlotDyadEstimates(result_diana, which = "connectivity", showUnpaired = TRUE)
```

This faceted view helps interpret the connectivity model in context:
a group may show high connectivity _probability_ among its pairs while still
having a large fraction of completely unpaired clones.

For full control over the plot, the underlying data frames
(`result_diana$connectivity`, `result_diana$distance`) can be used
with `ggplot2` directly.


## Approach: mixed vs clustered

By default `approach = "mixed"` fits a GLMM with crossed random intercepts for
the subject at each endpoint (`(1|from_SubjectId) + (1|to_SubjectId)`). If the
random-effects structure is too complex for the data, `approach = "clustered"`
fits a plain GLM instead and uses cluster-robust standard errors (clustered on
a sorted subject-pair identifier) via `marginaleffects`.

```{r clustered}
result_clustered <- ModelTCRDyads(
  seuratObj_TCR     = seuratObj_TCR,
  chains            = "TRB",
  groupColumn       = "Phenotype",
  distanceThreshold = 100,
  approach          = "clustered",
  communityMethod   = "threshold",
  subjectIdCol      = "SubjectId"
)

print(result_clustered$connectivity)
```

## Estimand: effects vs means

By default (`estimand = "effects"`), `ModelTCRDyads()` returns pairwise
contrasts between `pair_type` levels via `avg_comparisons()`. To instead
obtain predicted values (marginal means) for each level, set
`estimand = "means"`:

```{r means}
result_means <- ModelTCRDyads(
  seuratObj_TCR     = seuratObj_TCR,
  chains            = "TRB",
  groupColumn       = "Phenotype",
  distanceThreshold = 100,
  estimand          = "means",
  communityMethod   = "threshold",
  subjectIdCol      = "SubjectId"
)

PlotDyadEstimates(result_means, which = "connectivity")
```

## Session info

```{r sessionInfo}
sessionInfo()
```
