---
title: "RunTcrClustering: Single vs Paired Chain Outputs"
author: "tcrClustR"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RunTcrClustering: Single vs Paired Chain Outputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

This vignette demonstrates how to use `RunTcrClustering()` to cluster TCR sequences and write compact parquet tables for downstream analysis. The function supports:

- **Single-chain assays**: Outputs have columns `chain`, `v_gene`, `j_gene`, `CDR3`, `Cluster`, ...
- **Paired-chain assays**: Outputs have columns `chain_1`, `v_gene_1`, `j_gene_1`, `CDR3_1`, `chain_2`, `v_gene_2`, `j_gene_2`, `CDR3_2`, `Cluster`, ...

Unlike other clustering approaches, this function expects you to have already subset your data. It operates on a single Seurat/TCR object and writes one parquet file per assay.

## Key Features

- **Automatic data filtering** (default `filterInvalidClones = TRUE`): Removes clones with:
  - NA values in chain-specific columns
  - Concatenated values (containing commas or semicolons)
  - For single-chain: filters if the specific chain has issues
  - For multi-chain: only filters if ALL chains have issues (keeps clones with at least one valid chain)

- **Allele stripping** (default `stripAlleles = TRUE`): Removes allele suffixes like `*01` from V/J gene names
  - `TRBV7-9*01` → `TRBV7-9`

# Setup

```{r load_packages, message=FALSE}
library(tcrClustR)
library(arrow)
library(dplyr)
```

# Load Example Data

We'll create a simple synthetic TCR dataset with known clusters for demonstration purposes.

```{r create_synthetic_data}
# Create synthetic TCR metadata with clear clusters
set.seed(123)

# Cluster 1: TRBV7-9 with similar CDR3s
cluster1 <- data.frame(
  TRB_V = rep("TRBV7-9", 5),
  TRB_J = rep("TRBJ2-1", 5),
  TRB = c("CASSLGQAYEQYF", "CASSLGQAYEQYF", "CASSLGQAYEQYF", 
          "CASSLAQAYEQYF", "CASSLAQAYEQYF"),
  SubjectId = paste0("S", 1:5),
  stringsAsFactors = FALSE
)

# Cluster 2: TRBV6-4 with similar CDR3s  
cluster2 <- data.frame(
  TRB_V = rep("TRBV6-4", 4),
  TRB_J = rep("TRBJ1-1", 4),
  TRB = c("CASSYLDTEAFF", "CASSYLDTEAFF", "CASSYLDAEAFF", "CASSYLDAEAFF"),
  SubjectId = paste0("S", 6:9),
  stringsAsFactors = FALSE
)

# Singletons
singletons <- data.frame(
  TRB_V = c("TRBV5-1", "TRBV12-3"),
  TRB_J = c("TRBJ2-3", "TRBJ1-2"),
  TRB = c("CASSLGQAYEQYF", "CASSYLDTEAFF"),  # Reuse existing sequences to simplify
  SubjectId = c("S10", "S11"),
  stringsAsFactors = FALSE
)

# Combine into metadata
metadata <- rbind(cluster1, cluster2, singletons)
rownames(metadata) <- paste0("clone_", seq_len(nrow(metadata)))

# Create a minimal Seurat object
# (We just need a placeholder counts matrix - the TCR data is in metadata)
library(Seurat)
dummy_counts <- matrix(
  rnorm(nrow(metadata) * 10), 
  nrow = 10, 
  ncol = nrow(metadata),
  dimnames = list(paste0("gene_", 1:10), rownames(metadata))
)

seurat_obj <- CreateSeuratObject(
  counts = dummy_counts,
  meta.data = metadata
)

cat("Created synthetic metadata:\n")
cat("  Total clones:", nrow(metadata), "\n")
cat("  Expected cluster 1:", sum(metadata$TRB == "CASSLGQAYEQYF" | metadata$TRB == "CASSLAQAYEQYF"), "clones\n")
cat("  Expected cluster 2:", sum(metadata$TRB == "CASSYLDTEAFF" | metadata$TRB == "CASSYLDAEAFF"), "clones\n")
cat("  Expected singletons:", 2, "clones\n")
```

# Step 1: Compute TCR Distance Matrices

First, we need to run `RunTcrdist3()` to compute distance matrices. For this demo, we'll use a single-chain (TRB only) approach to keep it simple.

```{r run_tcrdist3, message=FALSE}
# Run tcrdist3 on TRB chain
tcr_obj <- RunTcrdist3(
  seuratObj = seurat_obj,
  chains = "TRB",
  multichain = FALSE,
  minimumClonesPerSubject = 1,  # Keep all clones
  rdsOutputPath = tempdir(),
  verbose = FALSE
)
```

```{r check_assays}
# Check what assays were created
available_assays <- SeuratObject::Assays(tcr_obj)
cat("Available assays:\n")
print(available_assays)
```

# Step 2: Cluster with RunTcrClustering

Now we'll use `RunTcrClustering()` to cluster the assays and write parquet files. This demonstrates the **single-chain output format**.

## Single-Chain Clustering (TRB only)

```{r cluster_single_chain}
# Create output directory
output_dir <- file.path(tempdir(), "tcr_clustering_demo")
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

# Cluster TRB assay
results <- RunTcrClustering(
  seuratObj = tcr_obj,
  assays = "TRB",  # Single-chain assay
  clusteringMethod = "DIANA",  # UPPERCASE required!
  dianaHeight = 20,
  clusterSizeThreshold = 2,  # Remove singletons
  outputDir = output_dir,
  outputPrefix = "TRB_demo"
)
```

```{r inspect_single_results}
# Look at the clustering results
cat("Single-chain clustering results:\n")
cat("  Clusters found:", length(unique(results_single$TRB$cluster_assignments)), "\n")
cat("  Clones after thresholding:", 
    sum(results_single$TRB$cluster_assignments != -1), "\n")
```

### Read and Inspect Single-Chain Output

The parquet file will have the **single-chain schema**: `chain`, `v_gene`, `j_gene`, `CDR3`, `Cluster`, `Cluster_Size`, `Clustering_Method`

```{r read_single_chain}
# Find the parquet file
parquet_file <- list.files(output_dir, pattern = "TRB_demo.*\\.parquet$", full.names = TRUE)[1]
cat("Reading:", basename(parquet_file), "\n\n")

# Read the parquet file
single_chain_data <- arrow::read_parquet(parquet_file)

cat("Single-chain output schema:\n")
cat("Columns:", paste(colnames(single_chain_data), collapse = ", "), "\n")
cat("Dimensions:", nrow(single_chain_data), "rows x", ncol(single_chain_data), "columns\n\n")

# Show first few rows
cat("First 5 rows:\n")
print(head(single_chain_data, 5))
```

```{r analyze_single_clusters}
library(dplyr)

# Analyze cluster distribution
cluster_summary <- single_chain_data %>%
  filter(Cluster != -1) %>%  # Exclude noise cluster
  group_by(Cluster) %>%
  summarize(
    n_clones = n(),
    unique_v_genes = n_distinct(v_gene, na.rm = TRUE),
    unique_j_genes = n_distinct(j_gene, na.rm = TRUE),
    example_cdr3 = first(CDR3)
  ) %>%
  arrange(desc(n_clones))

cat("\nCluster sizes (excluding noise/singletons):\n")
print(cluster_summary)
cat("\nTotal clusters:", nrow(cluster_summary), "\n")
```

# Demonstrating Paired-Chain Output (Optional)

To demonstrate the paired-chain schema, we would need to:
1. Add TRA chain data to our synthetic dataset
2. Run `RunTcrdist3()` with `multichain = TRUE`  
3. Cluster a paired assay like `TRA_TRB`

For this simple demo, we'll show what the paired schema looks like conceptually:

**Paired-chain schema** (e.g., from `TRA_TRB` assay):
- `chain_1`, `v_gene_1`, `j_gene_1`, `CDR3_1` (first chain)
- `chain_2`, `v_gene_2`, `j_gene_2`, `CDR3_2` (second chain)
- `Cluster`, `Cluster_Size_Threshold`, `Clustering_Method`

# Schema Comparison Summary

| Feature | Single-Chain | Paired-Chain |
|---------|-------------|--------------|
| Chain columns | `chain` | `chain_1`, `chain_2` |
| V gene columns | `v_gene` | `v_gene_1`, `v_gene_2` |
| J gene columns | `j_gene` | `j_gene_1`, `j_gene_2` |
| CDR3 columns | `CDR3` | `CDR3_1`, `CDR3_2` |
| Use case | Single-chain analysis (TRB-only, TRA-only) | Multi-chain analysis (TRA+TRB paired) |

# Session Info

```{r session_info}
sessionInfo()
```

# Cleanup

```{r cleanup, include=FALSE}
# Clean up temporary files
unlink(output_dir, recursive = TRUE)
```

```
```

```{r inspect_single_output}
# Check what was created
cat("\nParquet files created:\n")
print(results_single$parquet_files)

cat("\nSummary:\n")
print(results_single$summary)
```

### Read and Inspect Single-Chain Output

```{r read_single_chain_output}
# Read the parquet file
single_chain_data <- arrow::read_parquet(results_single$parquet_files[1])

cat("\nSingle-chain output schema:\n")
cat("Columns:", paste(colnames(single_chain_data), collapse = ", "), "\n")
cat("Dimensions:", nrow(single_chain_data), "rows x", ncol(single_chain_data), "columns\n")

# Show first few rows
cat("\nFirst 5 rows:\n")
print(head(single_chain_data, 5))
```

```{r analyze_single_clusters}
# Analyze cluster distribution
cluster_summary <- single_chain_data %>%
  filter(Cluster != "0") %>%
  group_by(Cluster) %>%
  summarize(
    n_clones = n(),
    unique_v_genes = n_distinct(v_gene, na.rm = TRUE),
    unique_j_genes = n_distinct(j_gene, na.rm = TRUE)
  ) %>%
  arrange(desc(n_clones))

cat("\nCluster sizes (excluding noise/singletons):\n")
print(cluster_summary)
cat("\nTotal clusters:", nrow(cluster_summary), "\n")
cat("Total clones in clusters:", sum(cluster_summary$n_clones), "\n")
cat("Noise/singleton clones:", sum(single_chain_data$Cluster == "0"), "\n")
```

## Example 2: Paired-Chain Clustering (TRA + TRB)

```{r cluster_paired_chain}
# Create output directory
output_dir_paired <- file.path(tempdir(), "paired_chain_example")
dir.create(output_dir_paired, showWarnings = FALSE, recursive = TRUE)

# Cluster paired TRA+TRB assay (if it exists)
paired_assays <- grep("TRA.*TRB|TRB.*TRA", available_assays, value = TRUE)
if (length(paired_assays) > 0) {
  cat("Found paired assay:", paired_assays[1], "\n")
  
  results_paired <- RunTcrClustering(
    seuratObj_TCR = tcr_obj,
    assaysToCluster = paired_assays[1],  # Multi-chain assay
    clusteringMethod = "DIANA",
    dianaHeight = 25,  # Slightly higher threshold for paired
    clusterSizeThreshold = 2,
    outputDir = output_dir_paired,
    maxChainsToRecord = 2,
    verbose = TRUE
  )
  
  cat("\nParquet files created:\n")
  print(results_paired$parquet_files)
  
  cat("\nSummary:\n")
  print(results_paired$summary)
} else {
  cat("No paired-chain assays found. This is expected if multichain=FALSE in RunTcrdist3.\n")
}
```

### Read and Inspect Paired-Chain Output

```{r read_paired_chain, eval=length(paired_assays) > 0}
if (length(paired_assays) > 0 && length(results_paired$parquet_files) > 0) {
  # Read the parquet file
  paired_chain_data <- arrow::read_parquet(results_paired$parquet_files[1])
  
  cat("\nPaired-chain output schema:\n")
  cat("Columns:", paste(colnames(paired_chain_data), collapse = ", "), "\n")
  cat("Dimensions:", nrow(paired_chain_data), "rows x", ncol(paired_chain_data), "columns\n")
  
  # Show first few rows
  cat("\nFirst 5 rows:\n")
  print(head(paired_chain_data, 5))
}
```

```{r analyze_paired_clusters, eval=length(paired_assays) > 0}
if (length(paired_assays) > 0 && length(results_paired$parquet_files) > 0) {
  # Analyze cluster distribution
  cluster_summary_paired <- paired_chain_data %>%
    filter(Cluster != "0") %>%
    group_by(Cluster) %>%
    summarize(
      n_clones = n(),
      unique_v_genes_1 = n_distinct(v_gene_1, na.rm = TRUE),
      unique_v_genes_2 = n_distinct(v_gene_2, na.rm = TRUE)
    ) %>%
    arrange(desc(n_clones))
  
  cat("\nPaired cluster sizes (excluding noise/singletons):\n")
  print(cluster_summary_paired)
  cat("\nTotal paired clusters:", nrow(cluster_summary_paired), "\n")
}
```

## Example 3: Cluster All Assays at Once

You can also cluster all assays in one call:

```{r cluster_all_assays}
# Create output directory
output_dir_all <- file.path(tempdir(), "all_assays_example")
dir.create(output_dir_all, showWarnings = FALSE, recursive = TRUE)

# Cluster all available assays
results_all <- RunTcrClustering(
  seuratObj_TCR = tcr_obj,
  assaysToCluster = NULL,  # NULL means all assays
  clusteringMethod = "DIANA",
  dianaHeight = 20,
  clusterSizeThreshold = 2,
  outputDir = output_dir_all,
  verbose = FALSE
)

cat("Total parquet files created:", length(results_all$parquet_files), "\n")
cat("\nSummary of all assays:\n")
print(results_all$summary)
```

# Schema Comparison

## Single-Chain Schema

Single-chain outputs (e.g., `TRB_cdr3`) have this structure:

| Column | Description |
|--------|-------------|
| `chain` | Chain identifier (TRA, TRB, TRG, or TRD) |
| `v_gene` | V gene segment |
| `j_gene` | J gene segment |
| `CDR3` | CDR3 amino acid sequence |
| `Cluster` | Cluster assignment (0 = noise/singleton) |
| `Cluster_Size_Threshold` | Minimum cluster size used |
| `Clustering_Method` | DIANA or Leiden |
| `param_*` | Clustering parameters (e.g., `param_cutHeight`) |

## Paired-Chain Schema

Paired-chain outputs (e.g., `TRA_TRB`) have this structure:

| Column | Description |
|--------|-------------|
| `chain_1` | First chain identifier |
| `v_gene_1` | First chain V gene |
| `j_gene_1` | First chain J gene |
| `CDR3_1` | First chain CDR3 sequence |
| `chain_2` | Second chain identifier |
| `v_gene_2` | Second chain V gene |
| `j_gene_2` | Second chain J gene |
| `CDR3_2` | Second chain CDR3 sequence |
| `Cluster` | Cluster assignment |
| `Cluster_Size_Threshold` | Minimum cluster size |
| `Clustering_Method` | DIANA or Leiden |
| `param_*` | Clustering parameters |

# Downstream Analysis Tips

## Joining Outputs

Since each assay creates a separate parquet file, you can read and combine them:

```{r combine_example, eval=FALSE}
# Read all parquet files from a directory
all_files <- list.files(output_dir_all, pattern = "\\.parquet$", full.names = TRUE)

# For single-chain files, you can rbind them
single_files <- grep("_single\\.parquet$", all_files, value = TRUE)
if (length(single_files) > 0) {
  combined_single <- do.call(rbind, lapply(single_files, arrow::read_parquet))
}

# Paired-chain files can be processed separately or joined by clone ID
```

## Filtering by Cluster Size

```{r filter_example, eval=FALSE}
# Read a parquet file and filter to large clusters only
data <- arrow::read_parquet(results_single$parquet_files[1])

large_clusters <- data %>%
  group_by(Cluster) %>%
  filter(n() >= 5, Cluster != "0") %>%
  ungroup()
```

## Finding Shared CDR3s Across Clusters

```{r shared_cdr3_example, eval=FALSE}
# Find CDR3 sequences that appear in multiple clusters
cdr3_sharing <- data %>%
  filter(Cluster != "0") %>%
  group_by(CDR3) %>%
  summarize(
    n_clusters = n_distinct(Cluster),
    clusters = paste(unique(Cluster), collapse = ",")
  ) %>%
  filter(n_clusters > 1)
```

# Summary

`RunTcrClustering()` provides a streamlined interface for TCR clustering:

**Key Benefits:**
- Clean separation: subset your data first, then cluster
- Compact output schemas optimized for downstream analysis
- Built-in filtering for invalid clones (NA, concatenated values)
- Optional allele stripping for normalized gene names
- Flexible clustering methods (DIANA or Leiden)
- Parquet format for efficient storage and reading

**Typical Workflow:**
1. Subset your Seurat object to the cells/conditions of interest
2. Run `RunTcrdist3()` to compute distance matrices
3. Run `RunTcrClustering()` to cluster and export parquet files
4. Read parquet files for downstream analysis (plotting, statistics, etc.)

# Session Info

```{r session_info}
sessionInfo()
```

# Cleanup

```{r cleanup, include=FALSE}
# Clean up temporary files
unlink(output_dir_single, recursive = TRUE)
unlink(output_dir_paired, recursive = TRUE)
unlink(output_dir_all, recursive = TRUE)
```
