---
title: "tcrClustR Workflow: TCR Clustering and Analysis"
author: "Bimber Lab"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{tcrClustR Workflow: TCR Clustering and Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6
)
```

## Introduction

The `tcrClustR` package provides tools for analyzing T-cell receptor (TCR) data by computing distance matrices between TCRs and clustering them into families. This vignette is the consolidated, primary guide and emphasizes the core production workflow:

RunTcrdist3() → RunTcrClustering() → JoinClusteringResults()

It demonstrates the complete workflow using example data from the Rhesus Immunome Reference Atlas (https://bimberlab.github.io/RIRA/). Secondary sections outline the functions chained within `RunTcrClustering()` and their purposes.

## Setup

First, we load the required libraries and set up the analysis parameters:

```{r libraries, message=FALSE, warning=FALSE}
library(tcrClustR)
library(dplyr)
library(Seurat)

# Set analysis parameters
chains <- c("TRA", "TRB")
cleanMetadata <- TRUE
summarizeClones <- TRUE
minimumClonesPerSubject <- 2
resolutionParameters <- c(0.1)
pcaComponents <- 50
usePCA <- TRUE
verbose <- FALSE
pythonExecutable <- NULL

# Create output directory
outputPath <- tempdir()
```

## Load Example Data

We'll use the example dataset included with the package:

```{r loadData}
# Load example Seurat object with TCR data
data_path <- system.file("extdata", "small_RIRA.rds", package = "tcrClustR")
seuratObj <- readRDS(data_path)

print("Seurat object summary:")
print(paste0("Number of cells: ", ncol(seuratObj)))
print(paste0("Number of features: ", nrow(seuratObj)))

# Check for required TCR columns
tcr_columns <- c("TRA", "TRB", "TRA_V", "TRA_J", "TRB_V", "TRB_J")
available_tcr_cols <- intersect(tcr_columns, colnames(seuratObj@meta.data))
print(paste0("Available TCR columns: ", paste(available_tcr_cols, collapse = ", ")))
```

## Primary Workflow

### Step 1: Format TCR Metadata

The first step is to format the TCR metadata for downstream analysis:

```{r formatMetadata}
formatted_metadata <- FormatMetadataForTcrDist3(
  metadata = seuratObj@meta.data,
  chains = chains,
  cleanMetadata = cleanMetadata,
  summarizeClones = summarizeClones,
  minimumClonesPerSubject = minimumClonesPerSubject,
  outputCsv = file.path(outputPath, 'tcrDist3Input.csv'),
  pythonExecutable = pythonExecutable,
  verbose = verbose
)

print("Formatted metadata summary:")
print(paste0("Number of rows: ", nrow(formatted_metadata)))
print(paste0("Number of columns: ", ncol(formatted_metadata)))
print(paste0("Chains included: ", paste(chains, collapse = ", ")))

# Display first few rows
knitr::kable(head(formatted_metadata, n = 5), caption = "First 5 rows of formatted metadata")
```

### Step 2: Compute TCR Distance Matrices

Next, we compute distance matrices between TCRs using tcrdist3:

```{r computeDistances}
seuratObj_TCR <- RunTcrdist3(
  seuratObj = seuratObj,
  chains = chains,
  cleanMetadata = cleanMetadata,
  minimumClonesPerSubject = minimumClonesPerSubject,
  rdsOutputPath = file.path(outputPath, "tcrdist3DistanceMatrices"),
  pythonExecutable = pythonExecutable,
  verbose = verbose
)

print("TCR distance matrices computed successfully!")
print(paste0("Available assays: ", paste(SeuratObject::Assays(seuratObj_TCR), collapse = ", ")))
print(paste0("Number of cells in TCR object: ", ncol(seuratObj_TCR)))
```

Optionally, you may start at the `RunTcrdist3` step and set `formatMetadata = TRUE`.

### Step 3: Cluster TCRs

Now we cluster the TCRs based on their similarity using DIANA (hierarchical) or Leiden (network-based) methods.

```{r clusterTCRs}
clustering_result <- ClusterTcrs(
  seuratObj_TCR = seuratObj_TCR,
  resolutionParameters = resolutionParameters,
  usePCA = usePCA,
  pcaComponents = pcaComponents
)

print("TCR clustering completed successfully!")

# The clustering result is a list containing clustered TCR objects
print(paste0("Number of clustered TCR objects: ", length(clustering_result)))

# Check clustering columns in the single-chain object
single_chain_obj <- clustering_result$singleChainSeuratObject
if (!is.null(single_chain_obj)) {
  cluster_cols <- grep("TcrClustR_", colnames(single_chain_obj@meta.data), value = TRUE)
  print(paste0("Clustering columns in single-chain object: ", paste(cluster_cols, collapse = ", ")))
  
  # Show sample clustering results
  if (length(cluster_cols) > 0) {
    print("Sample clustering assignments:")
    sample_data <- single_chain_obj@meta.data[1:5, c("cdr3_b_aa", "cdr3_b_aa", cluster_cols[1]), drop = FALSE]
    print(sample_data)
  }
}
```

### Step 4: Join Clustering Results to Original Seurat Object

The parquet files are created during the RunTcrClustering step. You control their location using the `outputDir` argument, which defaults to `"./tcrclustering_output/"`. Point `JoinClusteringResults()` at this same directory or pass the explicit parquet file paths returned from `RunTcrClustering()`.

```{r joinClustering}

#Use the packaged helper to join parquet clustering outputs back to the original Seurat object
library(tcrClustR)

#If you ran RunTcrClustering() earlier in this workflow and kept track of the parquet file paths:
#seurat_with_families <- JoinClusteringResults(
#  seuratObj = seuratObj,
#  parquetFiles = clustering_output$parquet_files,
#  metadataColumnPrefix = "TcrFamily",
#  stripAlleles = TRUE,
#  overwriteExisting = TRUE,
#  verbose = TRUE
#)

"#Alternatively, detect parquet files from the clustering output directory used by RunTcrClustering():"

#By default, RunTcrClustering writes parquet files to "./tcrclustering_output/".
#If you set a custom outputDir in RunTcrClustering(outputDir = "./my_output/"), use that here.
seurat_with_families <- JoinClusteringResults(
  seuratObj = seuratObj,
  parquetDir = "./tcrclustering_output/",
  filePattern = "\\.parquet$",
  metadataColumnPrefix = "TcrFamily",
  stripAlleles = TRUE,
  overwriteExisting = TRUE,
  verbose = TRUE
)

#Tip: if you ran RunTcrClustering() earlier in this vignette, you can also pass
#the explicit list of parquet files returned in clustering_output$parquet_files.

#Preview newly added metadata columns
new_cols <- grep("^TcrFamily_", colnames(seurat_with_families@meta.data), value = TRUE)
print(new_cols)
head(seurat_with_families@meta.data[, new_cols, drop = FALSE])
```

The `clustering_result` object contains clustered TCR Seurat objects:
- `singleChainSeuratObject` - TCRs clustered using individual chain distances (e.g. TRA, TRB, TRA_cdr3, TRB_cdr3). Each clone is indexed by subject + individual chain sequence.
- `multiChainSeuratObject` - TCRs clustered using combined multi-chain distances (e.g. TRA + TRB, TRA_cdr3 + TRB_cdr3). Each clone represents a paired TCR combination.

The clustering results are stored in the metadata of these objects with column names starting with "TcrClustR_". 
For single-chain results, the clonotypic join matches on subject + individual chain sequence (TRA or TRB).
For multi-chain results, the join would use the full TCR pair information.

## Explore Clustering Results

Now that we have clustering results joined to our original Seurat object, let's explore the results:

```{r exploreResults}

# Summary of cluster assignments
cluster_table <- table(seuratObj@meta.data[["TcrClustR_TRA_0.1"]], useNA = "always")
print(cluster_table)
```

## Visualize Results

Generate visualizations of the TCR clustering results using the clustered TCR objects:

```{r visualizations, fig.width=12, fig.height=8}
# Heatmaps - may have issues with current data structure
tryCatch({
  TCRDistanceHeatmaps(
    seuratObj_TCR = single_chain_obj,
    resolution = resolutionParameters[1],
    annotate_clusters = TRUE,
    verbose = verbose
  )
}, error = function(e) {
  cat("Error in TCRDistanceHeatmaps:", e$message, "\n")
  cat("This appears to be an internal bug in the heatmap annotation function.\n")
  cat("Skipping heatmap generation.\n")
})

# UMAP and PCA plots
umap_reduction <- "TcrClustR_pca.TRA.cdr3_umap"
if (umap_reduction %in% names(single_chain_obj@reductions)) {
  print(DimPlot(single_chain_obj,
          reduction = umap_reduction,
          group.by = "TcrClustR_TRA_cdr3_0.1",
          label = TRUE,
          pt.size = 0.5
        ))
} else {
  cat("UMAP reduction", umap_reduction, "not found\n")
}

# PCA reduction
pca_reduction <- "TcrClustR_pca.TRA.cdr3"
if (pca_reduction %in% names(single_chain_obj@reductions)) {
  print(DimPlot(
          single_chain_obj,
          reduction = pca_reduction,
          group.by = "TcrClustR_TRA_cdr3_0.1",
          label = TRUE,
          pt.size = 0.5
        ))
} else {
  cat("PCA reduction", pca_reduction, "not found\n")
}
```

## Understanding the Results

The tcrClustR workflow produces several key outputs:

1. **Distance Matrices**: Pairwise distances between TCRs based on various metrics
2. **Clustering Results**: TCRs grouped into clusters based on similarity
3. **Visualizations**: Heatmaps and histograms showing clustering patterns
4. **Clonotypic Join**: Transfer of clustering results back to the original Seurat object

### Interpreting Clusters

- Each cluster represents a group of similar TCRs that may have related functions
- Cluster assignments are added as new columns to your Seurat object metadata after the clonotypic join
- Cells without TCR data will have `NA` values for clustering columns

### Next Steps

With clustering results now available in your main Seurat object, you can:

1. **Analyze cluster composition**: Examine which cell types or conditions are enriched in each TCR cluster
2. **Functional analysis**: Investigate whether TCR clusters correlate with specific cellular functions
3. **Comparative studies**: Compare TCR clustering patterns across different experimental conditions
4. **Integration with other data**: Combine TCR clustering with gene expression, surface marker, or other omics data
- The resolution parameter controls the granularity of clustering (lower = fewer, larger clusters)

### Next Steps

After clustering, you can:

- Extract specific clusters for further analysis
- Use cluster assignments for downstream functional studies
- Prioritize representative TCRs from each cluster for synthesis
- Correlate clusters with phenotypic or functional data

## Output Files

```{r outputFiles}
cat("Files created in output directory:\n")

files <- list.files(outputPath, recursive = TRUE)
for (file in files) {
  cat("  ", file, "\n")
}
```

## Session Information

```{r sessionInfo}
sessionInfo()
```

## Conclusion

This vignette demonstrated the complete tcrClustR workflow for analyzing TCR data. The package provides a streamlined approach to:

1. Format TCR metadata for analysis
2. Compute distance matrices between TCRs
3. Cluster TCRs based on similarity
4. Visualize and interpret results

The resulting clusters can be used to identify groups of potentially functionally related TCRs for downstream analysis and experimental validation.

---

## Secondary: What `RunTcrClustering()` Does Under The Hood

This section briefly illustrates the purpose of the functions chained within `RunTcrClustering()`:

- `.filter_clones_for_assay()`
  - Input: formatted metadata; target assay chains
  - Output: filtered set of clones for the assay
  - Logic: strict filtering for single-chain assays (all columns valid), permissive for multi-chain (at least one chain valid); detects concatenations ("," or ";").

- `.strip_allele_suffix()`
  - Input: gene names possibly containing allele notation (e.g., `TRBV7-9*01`)
  - Output: normalized gene names without allele suffix

- `.DistanceMatrixToClusteredGraphs()` and `.PcaAndClustering()`
  - Input: distance matrices stored as Seurat assays
  - Output: cluster assignments based on PCA/KPCA reductions and kNN graphs; supports jaccard, rbf, and jaccard×rbf edge weighting; clustering via DIANA or Leiden.

- `.ComputeMultiTCRDistanceMatrix()`
  - Input: single-chain distance matrices
  - Output: combined multi-chain matrices for paired analysis (e.g., `TRA_TRB`).

- `WriteClusteringResultsToParquet()`
  - Input: cluster assignments and associated metadata
  - Output: compact parquet files with the appropriate schema for single vs paired TCRs.

These utilities are orchestrated by `RunTcrClustering()` to provide a streamlined interface with sensible defaults for end users.
