---
title: "tcrClustR Workflow: TCR Clustering and Analysis"
author: "GW McElfresh, Sebastian Benjamin, & Bimber Lab"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{tcrClustR Workflow: TCR Clustering and Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6
)
```

## Introduction

The `tcrClustR` package provides tools for analyzing T-cell receptor (TCR) data by computing distance matrices between TCRs and clustering them into families. This vignette is the consolidated, primary guide and emphasizes the core production workflow:

RunTcrdist3() → RunTcrClustering() → JoinClusteringResults()

It demonstrates the complete workflow using example data from the Rhesus Immunome Reference Atlas (https://bimberlab.github.io/RIRA/). Secondary sections outline the functions chained within `RunTcrClustering()` and their purposes.

## Setup

First, we load the required libraries and set up the analysis parameters:

```{r libraries, message=FALSE, warning=FALSE}
library(tcrClustR)
library(dplyr)
library(Seurat)
library(arrow)

# Set analysis parameters
chains <- c("TRA", "TRB")
summarizeClones <- TRUE
minimumClonesPerSubject <- 2
resolutionParameters <- c(0.1)
pcaComponents <- 50
usePCA <- TRUE
verbose <- FALSE

# Create output directory
outputPath <- tempdir()
```

## Load Example Data

We'll use the example dataset included with the package:

```{r loadData}
# Load example Seurat object with TCR data
data_path <- system.file("extdata", "small_RIRA.rds", package = "tcrClustR")
seuratObj <- readRDS(data_path)

print("Seurat object summary:")
print(paste0("Number of cells: ", ncol(seuratObj)))
print(paste0("Number of features: ", nrow(seuratObj)))

# Check for required TCR columns
tcr_columns <- c("TRA", "TRB", "TRA_V", "TRA_J", "TRB_V", "TRB_J")
available_tcr_cols <- intersect(tcr_columns, colnames(seuratObj@meta.data))
print(paste0("Available TCR columns: ", paste(available_tcr_cols, collapse = ", ")))
```

## Primary Workflow

### 1: Format TCR Metadata (Optional)

The first step is to format the TCR metadata for downstream analysis:

```{r formatMetadata}
formatted_metadata <- FormatMetadataForTcrDist3(
  metadata = seuratObj@meta.data,
  chains = chains,
  summarizeClones = summarizeClones,
  minimumClonesPerSubject = minimumClonesPerSubject,
  outputPath = outputPath,
  verbose = verbose
)

print("Formatted metadata summary:")
print(paste0("Number of rows: ", nrow(formatted_metadata)))
print(paste0("Number of columns: ", ncol(formatted_metadata)))
print(paste0("Chains included: ", paste(chains, collapse = ", ")))

# Display first few rows
knitr::kable(head(formatted_metadata, n = 5), caption = "First 5 rows of formatted metadata")
```

### 2: Compute TCR Distance Matrices

Note: You may start at the `RunTcrdist3` step.

Next, we compute distance matrices between TCRs using tcrdist3:



```{r computeDistances}
seuratObj_TCR <- RunTcrdist3(
  inputData = seuratObj,
  chains = chains,
  minimumClonesPerSubject = minimumClonesPerSubject,
  rdsOutputPath = file.path(outputPath, "tcrdist3DistanceMatrices"),
  verbose = verbose
)

print("TCR distance matrices computed successfully!")
print(paste0("Available assays: ", paste(SeuratObject::Assays(seuratObj_TCR), collapse = ", ")))
print(paste0("Number of cells in TCR object: ", ncol(seuratObj_TCR)))
```


### 3: Cluster TCRs

Now we cluster the TCRs based on their similarity using DIANA (hierarchical) or Leiden (network-based) methods.

```{r clusterTCRs}
clustering_result <- RunTcrClustering(
  seuratObj_TCR = seuratObj_TCR,
  assaysToCluster = "TRB",
  resolutionParameters = resolutionParameters,
  usePCA = usePCA,
  dianaHeight = 20, 
  clusterSizeThreshold = 1,
  clusteringMethod = "DIANA",
  pcaComponents = pcaComponents, 
  outputDir = file.path(outputPath, "tcrclustering_output"),
  verbose = verbose, 
  stripAlleles = TRUE
)

print("TCR clustering completed successfully!")

# View intermediate TCR Clustering files
clustering_result$parquet_files
parquet_summary <- head(arrow::read_parquet(clustering_result$parquet_files[[1]]))
print("Preview of clustering parquet file:")
knitr::kable(parquet_summary, caption = "First 5 rows of clustering parquet file")

```

### 4: Join Clustering Results to Original Seurat Object

The parquet files are created during the RunTcrClustering step. You control their location using the `outputDir` argument, which defaults to `"./tcrclustering_output/"`. Point `JoinClusteringResults()` at this same directory or pass the explicit parquet file paths returned from `RunTcrClustering()`.

```{r joinClustering}

#Use the packaged helper to join parquet clustering outputs back to the original Seurat object
library(tcrClustR)

#By default, RunTcrClustering writes parquet files to "./tcrclustering_output/".
#If you set a custom outputDir in RunTcrClustering(outputDir = "./my_output/"), use that here.
seurat_with_families <- JoinClusteringResults(
  seuratObj = seuratObj,
  parquetDir = file.path(outputPath, "tcrclustering_output"),
  filePattern = "\\.parquet$",
  metadataColumnPrefix = "TcrFamily",
  stripAlleles = TRUE,
  overwriteExisting = TRUE,
  verbose = TRUE
)


#Preview newly added metadata columns
new_cols <- grep("^TcrFamily_", colnames(seurat_with_families@meta.data), value = TRUE)
print(new_cols)

# Clusters are automatically formatted as factors with levels:
# - Numeric clusters (1, 2, 3, ...) sorted first
# - "LowFrequency" for filtered/singleton clones
# - "No_TCR_Data" for cells without TCR information
print("Cluster levels:")
print(levels(seurat_with_families@meta.data[[new_cols[1]]]))

print("Cluster distribution:")
print(table(seurat_with_families@meta.data[[new_cols[1]]], useNA = "always"))
```

```{r}
#Visualize TCR families on UMAP
# Note: Clusters are already formatted as factors by JoinClusteringResults
# Values: "1", "2", "3", ... (numeric clusters), "LowFrequency", "No_TCR_Data"

#check that you have a UMAP-able number of classes: 
unique_classes <- length(levels(seurat_with_families@meta.data[[new_cols[1]]]))
print(paste0("Number of unique TCR families (including LowFrequency and No_TCR_Data): ", unique_classes))

if (unique_classes > 25) {
  stop("Too many unique TCR families for effective UMAP visualization. Consider subsetting or using a different visualization approach.")
}

#Visualize
DimPlot(
  seurat_with_families,
  reduction = "umap",
  group.by = new_cols,
  label = TRUE,
  pt.size = 1
) 
```


The `clustering_result` object contains:
- `parquet_files` - Vector of file paths to the parquet files written to disk
- `summary` - Data frame with clustering summary per assay (number of clones, clusters, etc.)

These parquet files contain the clustering assignments and can be joined back to your original Seurat object using `JoinClusteringResults()` (shown in step 4 above).

## Explore Clustering Results

After joining clustering results back to the original Seurat object (step 4), explore the cluster assignments:

```{r exploreResults}
# Summary of cluster assignments (from the joined metadata columns)
# The column name depends on the metadataColumnPrefix used in JoinClusteringResults()
cluster_cols <- grep("^TcrFamily_", colnames(seurat_with_families@meta.data), value = TRUE)

if (length(cluster_cols) > 0) {
  for (col in cluster_cols) {
    cat("\nCluster distribution for", col, ":\n")
    print(table(seurat_with_families@meta.data[[col]], useNA = "always"))
  }
} else {
  message("No TcrFamily cluster columns found. Run JoinClusteringResults() first.")
}
```

## Visualize TCR Distance Matrices

After computing distance matrices with `RunTcrdist3()`, you can visualize the pairwise distances between TCR clones. The distance matrices are stored as Seurat assays in `seuratObj_TCR`.

### Basic Distance Heatmap

Create a simple heatmap showing TCR distances with hierarchical clustering:

```{r distanceHeatmap, fig.width=8, fig.height=8}
library(ComplexHeatmap)

# Extract distance matrix for TRB assay
# Available assays depend on what chains were included in RunTcrdist3()
distance_matrix <- Seurat::GetAssayData(
  seuratObj_TCR, 
  assay = "TRB",  # Use "TRA", "TRB", "TRA_cdr3", or "TRB_cdr3" depending on your analysis
  layer = "counts"
)

# Create a hierarchical clustered heatmap
distance_heatmap <- ComplexHeatmap::Heatmap(
  as.matrix(distance_matrix),
  name = "TCR Distance",
  show_row_names = FALSE,
  show_column_names = FALSE,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  clustering_method_rows = "ward.D2",
  clustering_method_columns = "ward.D2",
  use_raster = TRUE,
  show_heatmap_legend = TRUE,
  column_title = "TRB Distance Matrix"
)

draw(distance_heatmap)
```

### Distance Histogram

Visualize the distribution of pairwise TCR distances:

```{r distanceHistogram, fig.width=8, fig.height=5}
# Extract upper triangle of distance matrix (avoid duplicates)
dist_values <- distance_matrix[upper.tri(distance_matrix)]

# Plot histogram
hist(
  dist_values,
  breaks = 50,
  main = "Distribution of TCR Pairwise Distances",
  xlab = "Distance",
  ylab = "Frequency",
  col = "steelblue",
  border = "white"
)
```

### Using TCRDistanceHeatmaps() with ClusterTcrs()

**Note:** The built-in `TCRDistanceHeatmaps()` function requires clustering metadata columns (e.g., `TcrClustR_TRB_0.1`) which are added by the `ClusterTcrs()` function, not by `RunTcrClustering()`. 

If you want to use `TCRDistanceHeatmaps()`, use `ClusterTcrs()` instead of `RunTcrClustering()`:

```{r clusterTcrsExample, eval=FALSE}
# Alternative workflow using ClusterTcrs() instead of RunTcrClustering()
# This adds clustering metadata directly to the Seurat object
seuratObj_clustered <- ClusterTcrs(
  seuratObj_TCR = seuratObj_TCR,
  resolutionParameters = c(0.1),
  usePCA = TRUE,
  pcaComponents = 50,
  verbose = TRUE
)


heatmap <- TCRDistanceHeatmaps(
  seuratObj_TCR = seuratObj_clustered$singleChainSeuratObject,
  assayList = c("TRB"),
  annotate_clusters = FALSE,
  verbose = FALSE
)
heatmap
```

The primary workflow demonstrated in this vignette (`RunTcrClustering()`) focuses on creating compact parquet output files for downstream analysis, while `ClusterTcrs()` is better suited for interactive exploration with built-in visualization functions.

## Understanding the Results

The tcrClustR workflow produces several key outputs:

1. **Distance Matrices**: Pairwise distances between TCRs based on various metrics
2. **Clustering Results**: TCRs grouped into clusters based on similarity
3. **Visualizations**: Heatmaps and histograms showing clustering patterns
4. **Clonotypic Join**: Transfer of clustering results back to the original Seurat object

### Interpreting Clusters

- Each cluster represents a group of similar TCRs that may have related functions
- Cluster assignments are added as new columns to your Seurat object metadata after the clonotypic join
- Cells without TCR data will have `NA` values for clustering columns

### Next Steps

With clustering results now available in your main Seurat object, you can:

1. **Analyze cluster composition**: Examine which cell types or conditions are enriched in each TCR cluster
2. **Functional analysis**: Investigate whether TCR clusters correlate with specific cellular functions
3. **Comparative studies**: Compare TCR clustering patterns across different experimental conditions
4. **Integration with other data**: Combine TCR clustering with gene expression, surface marker, or other omics data
- The resolution parameter controls the granularity of clustering (lower = fewer, larger clusters)

### Next Steps

After clustering, you can:

- Extract specific clusters for further analysis
- Use cluster assignments for downstream functional studies
- Prioritize representative TCRs from each cluster for synthesis
- Correlate clusters with phenotypic or functional data

## Output Files

```{r outputFiles}
cat("Files created in output directory:\n")

files <- list.files(outputPath, recursive = TRUE)
for (file in files) {
  cat("  ", file, "\n")
}
```

## Session Information

```{r sessionInfo}
sessionInfo()
```

## Conclusion

This vignette demonstrated the complete tcrClustR workflow for analyzing TCR data. The package provides a streamlined approach to:

1. Format TCR metadata for analysis
2. Compute distance matrices between TCRs
3. Cluster TCRs based on similarity
4. Visualize and interpret results

The resulting clusters can be used to identify groups of potentially functionally related TCRs for downstream analysis and experimental validation.

---

## Secondary: What `RunTcrClustering()` Does Under The Hood

This section briefly illustrates the purpose of the functions chained within `RunTcrClustering()`:

- `.filter_clones_for_assay()`
  - Input: formatted metadata; target assay chains
  - Output: filtered set of clones for the assay
  - Logic: strict filtering for single-chain assays (all columns valid), permissive for multi-chain (at least one chain valid); detects concatenations ("," or ";").

- `.strip_allele_suffix()`
  - Input: gene names possibly containing allele notation (e.g., `TRBV7-9*01`)
  - Output: normalized gene names without allele suffix

- `.DistanceMatrixToClusteredGraphs()` and `.PcaAndClustering()`
  - Input: distance matrices stored as Seurat assays
  - Output: cluster assignments based on PCA/KPCA reductions and kNN graphs; supports jaccard, rbf, and jaccard×rbf edge weighting; clustering via DIANA or Leiden.

- `.ComputeMultiTCRDistanceMatrix()`
  - Input: single-chain distance matrices
  - Output: combined multi-chain matrices for paired analysis (e.g., `TRA_TRB`).

- `WriteClusteringResultsToParquet()`
  - Input: cluster assignments and associated metadata
  - Output: compact parquet files with the appropriate schema for single vs paired TCRs.

These utilities are orchestrated by `RunTcrClustering()` to provide a streamlined interface with sensible defaults for end users.
