---
title: "tcrClustR Workflow: TCR Clustering and Analysis"
author: "GW McElfresh, Sebastian Benjamin, & Bimber Lab"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{tcrClustR Workflow: TCR Clustering and Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6
)
```

## Introduction

The `tcrClustR` package provides tools for analyzing T-cell receptor (TCR) data by computing distance matrices between TCRs and clustering them into families. This vignette is the consolidated, primary guide and emphasizes the core production workflow:

CalculateTcrDistances() → RunTcrClustering()

It demonstrates the complete workflow using example data from the Rhesus Immunome Reference Atlas (https://bimberlab.github.io/RIRA/). Secondary sections outline the functions chained within `RunTcrClustering()` and their purposes.

## Setup

First, we load the required libraries and set up the analysis parameters:

```{r libraries, message=FALSE, warning=FALSE}

library(tcrClustR)
library(Seurat)

```

## Load Example Data

We'll use the example dataset included with the package:

```{r loadData}
# Load example Seurat object with TCR data
data_path <- system.file("extdata", "small_RIRA.rds", package = "tcrClustR")
seuratObj <- readRDS(data_path)

print("Seurat object summary:")
print(paste0("Number of cells: ", ncol(seuratObj)))
print(paste0("Number of features: ", nrow(seuratObj)))

# Check for required TCR columns
tcr_columns <- c("TRA", "TRB", "TRA_V", "TRA_J", "TRB_V", "TRB_J")
available_tcr_cols <- intersect(tcr_columns, colnames(seuratObj@meta.data))
print(paste0("Available TCR columns: ", paste(available_tcr_cols, collapse = ", ")))
```

## Primary Workflow


### 1: Compute TCR Distance Matrices

This will take the TCR CDR3/V/J information (either stored in the seuratObj metadata or as a dataframe) and compute pairwise distances using tcrdist3:

```{r computeDistances}

seuratObj_TCR <- CalculateTcrDistances(
  inputData = seuratObj,
  chains = c("TRA", "TRB"),
  minimumCloneSize = 2,
  calculateChainPairs = TRUE
)

print("TCR distance matrices computed successfully!")
print(paste0("Available assays: ", paste(SeuratObject::Assays(seuratObj_TCR), collapse = ", ")))
print(paste0("Number of cells in TCR object: ", ncol(seuratObj_TCR)))
```


### 2: Cluster TCRs

Now we cluster the TCRs based on their similarity using DIANA (hierarchical) clustering. The resulting seurat object contains the raw distance matrices under the @misc slot. Cells are assigned to a family/cluster index for each chain:

```{r clusterTCRs}

seuratObj_TCR <- RunTcrClustering(
  seuratObj_TCR = seuratObj_TCR,
  dianaHeight = 20, 
  clusterSizeThreshold = 1
)

print("TCR clustering completed successfully!")
```


```{r}

# Visualize
DimPlot(
  seuratObj_TCR,
  reduction = "umap",
  group.by = 'TRB_ClusterIdx',
  label = TRUE,
  pt.size = 1
)
```

## Visualize TCR Distance Matrices

After computing distance matrices with `CalculateTcrDistances()`, you can visualize the pairwise distances between TCR clones. The distance matrices are stored as Seurat assays in `seuratObj_TCR`.

### Basic Distance Heatmap

Create a simple heatmap showing TCR distances with hierarchical clustering:

```{r distanceHeatmap, fig.width=8, fig.height=8}
library(ComplexHeatmap)

# Extract distance matrix for TRB assay
distance_matrix <- GetDistanceMatrix(seuratObj_TCR, chains = "TRB")

# Create a hierarchical clustered heatmap
distance_heatmap <- ComplexHeatmap::Heatmap(
  as.matrix(distance_matrix),
  name = "TCR Distance",
  show_row_names = FALSE,
  show_column_names = FALSE,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  clustering_method_rows = "ward.D2",
  clustering_method_columns = "ward.D2",
  use_raster = TRUE,
  show_heatmap_legend = TRUE,
  column_title = "TRB Distance Matrix"
)

draw(distance_heatmap)
```

### Distance Histogram

Visualize the distribution of pairwise TCR distances:

```{r distanceHistogram, fig.width=8, fig.height=5}
# Extract upper triangle of distance matrix (avoid duplicates)
dist_values <- distance_matrix[upper.tri(distance_matrix)]

# Plot histogram
hist(
  dist_values,
  breaks = 50,
  main = "Distribution of TCR Pairwise Distances",
  xlab = "Distance",
  ylab = "Frequency",
  col = "steelblue",
  border = "white"
)
```

## Understanding the Results

The tcrClustR workflow produces several key outputs:

1. **Distance Matrices**: Pairwise distances between TCRs based on various metrics
2. **Clustering Results**: TCRs grouped into clusters based on similarity
3. **Visualizations**: Heatmaps and histograms showing clustering patterns
4. **Clonotypic Join**: Transfer of clustering results back to the original Seurat object

### Interpreting Clusters

- Each cluster represents a group of similar TCRs that may have related functions
- Cluster assignments are added as new columns to your Seurat object metadata after the clonotypic join
- Cells without TCR data will have `NA` values for clustering columns

### Next Steps

With clustering results now available in your main Seurat object, you can:

1. **Analyze cluster composition**: Examine which cell types or conditions are enriched in each TCR cluster
2. **Functional analysis**: Investigate whether TCR clusters correlate with specific cellular functions
3. **Comparative studies**: Compare TCR clustering patterns across different experimental conditions
4. **Integration with other data**: Combine TCR clustering with gene expression, surface marker, or other omics data
- The resolution parameter controls the granularity of clustering (lower = fewer, larger clusters)

### Next Steps

After clustering, you can:

- Extract specific clusters for further analysis
- Use cluster assignments for downstream functional studies
- Prioritize representative TCRs from each cluster for synthesis
- Correlate clusters with phenotypic or functional data


## Session Information

```{r sessionInfo}
sessionInfo()
```

## Conclusion

This vignette demonstrated the complete tcrClustR workflow for analyzing TCR data. The package provides a streamlined approach to:

1. Format TCR metadata for analysis
2. Compute distance matrices between TCRs
3. Cluster TCRs based on similarity
4. Visualize and interpret results

The resulting clusters can be used to identify groups of potentially functionally related TCRs for downstream analysis and experimental validation.